---
title: "pbmc_pipeline"
author: "Theresa Alexander"
date: "`r Sys.Date()`"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages
```{r}
library(ggplot2)
library(devtools)
library(scran)
library(iDA)
library(Seurat)
library(clValid)
library(caret)
library(Rtsne)
library(umap)
```


# Load the PBMC dataset
The 3k cell PBMC dataset can be downloaded from here:

https://cf.10xgenomics.com/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz"

```{r}
pbmc.data <- Read10X(data.dir = "~/Desktop/iDA_bioc/data/filtered_gene_bc_matrices/hg19/")
# Initialize the Seurat object with the raw (non-normalized data).
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "pbmc3k", min.cells = 3, min.features = 200)

pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
pbmc <- NormalizeData(pbmc, normalization.method = "LogNormalize", scale.factor = 10000)
#all.genes <- rownames(pbmc)
pbmc <- ScaleData(pbmc)
```

# iDA
```{r}
#run iDA
pbmc <- iDA(pbmc, assay = "RNA", set.seed = 11, diag = TRUE)
```


# PCA
```{r}
pbmc <- RunPCA(pbmc, features = VariableFeatures(pbmc))
```

```{r}
#iDA embedding and idents
pbmc_iDA_ident = pbmc@meta.data$iDA_clust
pbmc_iDAcellembedding = pbmc[["iDA"]]@cell.embeddings
```

```{r}
#embedding
pbmc_PCAembedding = Embeddings(pbmc, reduction = "pca")[, 1:10]
```


#cumulative F-Statistics and R^2 for iDA and PCA embeddings

##iDA F-stat and R^2
```{r}

#getting RSS for iDA across dimensions
iDARSS = c()
for (i in 1:dim(pbmc_iDAcellembedding)[2]) {
  iDARSS[i] = anova(lm(as.numeric(pbmc_iDAcellembedding[,i]) ~ as.factor(pbmc_iDA_ident)))[2,2]
}



#getting TSS for iDA across dimensions
iDATSS = c()
for (i in 1:dim(pbmc_iDAcellembedding)[2]) {
  iDATSS[i] = sum(anova(lm(as.numeric(pbmc_iDAcellembedding[,i]) ~ as.factor(pbmc_iDA_ident)))[,2])
}

Fstat_pbmc_iDA = c((iDATSS[1]-iDARSS[1])/iDARSS[1], 
                       (sum(iDATSS[1:2])-sum(iDARSS[1:2]))/sum(iDARSS[1:2]), 
                       (sum(iDATSS[1:3])-sum(iDARSS[1:3]))/sum(iDARSS[1:3]), 
                       (sum(iDATSS[1:4])-sum(iDARSS[1:4]))/sum(iDARSS[1:4]), 
                       (sum(iDATSS[1:5])-sum(iDARSS[1:5]))/sum(iDARSS[1:5]), 
                       (sum(iDATSS[1:6])-sum(iDARSS[1:6]))/sum(iDARSS[1:6]),   
                       (sum(iDATSS[1:7])-sum(iDARSS[1:7]))/sum(iDARSS[1:7]),  
                       (sum(iDATSS[1:8])-sum(iDARSS[1:8]))/sum(iDARSS[1:8]),
                       (sum(iDATSS[1:9])-sum(iDARSS[1:9]))/sum(iDARSS[1:9]),
                       (sum(iDATSS[1:10])-sum(iDARSS[1:10]))/sum(iDARSS[1:10]))

R2_pbmc_iDA = c(1-(iDARSS[1]/iDATSS[1]), 
                       1-sum(iDARSS[1:2])/sum(iDATSS[1:2]), 
                       1-sum(iDARSS[1:3])/sum(iDATSS[1:3]), 
                       1-sum(iDARSS[1:4])/sum(iDATSS[1:4]), 
                       1-sum(iDARSS[1:5])/sum(iDATSS[1:5]), 
                       1-sum(iDARSS[1:6])/sum(iDATSS[1:6]),   
                       1-sum(iDARSS[1:7])/sum(iDATSS[1:7]),  
                       1-sum(iDARSS[1:8])/sum(iDATSS[1:8]),  
                       1-sum(iDARSS[1:9])/sum(iDATSS[1:9]),  
                       1-sum(iDARSS[1:10])/sum(iDATSS[1:10]))


```

## PCA F-stat and R^2

```{r}
#getting RSS for iDA across dimensions
PCRSS = c()
for (i in 1:10) {
  PCRSS[i] = anova(lm(as.numeric(pbmc_PCAembedding[,i]) ~ as.factor(pbmc_iDA_ident)))[2,2]
}


#getting TSS for iDA across dimensions
PCTSS = c()
for (i in 1:10) {
  PCTSS[i] = sum(anova(lm(as.numeric(pbmc_PCAembedding[,i]) ~ as.factor(pbmc_iDA_ident)))[,2])
}


Fstat_pbmc_PCA = c((PCTSS[1]-PCRSS[1])/PCRSS[1], 
                       (sum(PCTSS[1:2])-sum(PCRSS[1:2]))/sum(PCRSS[1:2]), 
                       (sum(PCTSS[1:3])-sum(PCRSS[1:3]))/sum(PCRSS[1:3]), 
                       (sum(PCTSS[1:4])-sum(PCRSS[1:4]))/sum(PCRSS[1:4]), 
                       (sum(PCTSS[1:5])-sum(PCRSS[1:5]))/sum(PCRSS[1:5]), 
                       (sum(PCTSS[1:6])-sum(PCRSS[1:6]))/sum(PCRSS[1:6]),   
                       (sum(PCTSS[1:7])-sum(PCRSS[1:7]))/sum(PCRSS[1:7]),  
                       (sum(PCTSS[1:8])-sum(PCRSS[1:8]))/sum(PCRSS[1:8]),  
                       (sum(PCTSS[1:9])-sum(PCRSS[1:9]))/sum(PCRSS[1:9]),  
                       (sum(PCTSS[1:10])-sum(PCRSS[1:10]))/sum(PCRSS[1:10]))


R2_pbmc_PCA = c(1-(PCRSS[1]/PCTSS[1]), 
                       1-sum(PCRSS[1:2])/sum(PCTSS[1:2]), 
                       1-sum(PCRSS[1:3])/sum(PCTSS[1:3]), 
                       1-sum(PCRSS[1:4])/sum(PCTSS[1:4]), 
                       1-sum(PCRSS[1:5])/sum(PCTSS[1:5]), 
                       1-sum(PCRSS[1:6])/sum(PCTSS[1:6]),   
                       1-sum(PCRSS[1:7])/sum(PCTSS[1:7]),  
                       1-sum(PCRSS[1:8])/sum(PCTSS[1:8]),  
                       1-sum(PCRSS[1:9])/sum(PCTSS[1:9]),  
                       1-sum(PCRSS[1:10])/sum(PCTSS[1:10]))

Fstat_pbmc = as.data.frame(cbind(Fstat_pbmc_iDA, Fstat_pbmc_PCA))
R2_pbmc = as.data.frame(cbind(R2_pbmc_iDA, R2_pbmc_PCA))
```


## plot F-stat and R^2
```{r}
ggplot(Fstat_pbmc, aes(x = as.factor(1:10))) + geom_point(aes(y = Fstat_pbmc_iDA,color = "iDA")) + geom_line(aes(x = 1:10, y = Fstat_pbmc_iDA, color = "iDA")) + geom_point(aes(y = Fstat_pbmc_PCA,color = "PCA")) + geom_line(aes(x = 1:10, y = Fstat_pbmc_PCA, color = "PCA"))  +  scale_colour_manual("", breaks = c("iDA", "PCA"), values = c("blue", "orange")) + labs( x = "Dimension", y = "Cumulative F-statistic") + theme_classic()

ggplot(R2_pbmc, aes(x = as.factor(1:10))) + geom_point(aes(y = R2_pbmc_iDA,color = "iDA")) + geom_line(aes(x = 1:10, y = R2_pbmc_iDA, color = "iDA")) + geom_point(aes(y = R2_pbmc_PCA,color = "PCA")) + geom_line(aes(x = 1:10, y = R2_pbmc_PCA, color = "PCA"))  +  scale_colour_manual("", breaks = c("iDA", "PCA"), values = c("blue", "orange")) + labs( x = "Dimension", y = "Cumulative" ~ R^2) + theme_classic()  

```

# Do LD's separate clusters by expression of known markers?
```{r}

#markers CD14/LYZ: Monocytes
CD14iDA = as.data.frame(cbind(pbmc_iDAcellembedding, pbmc[["RNA"]]@counts["CD14",], pbmc_iDA_ident))
LYZiDA = as.data.frame(cbind(pbmc_iDAcellembedding, pbmc[["RNA"]]@counts["LYZ",], pbmc_iDA_ident))

#marker MS4A1: Bcells
MS4A1iDA = as.data.frame(cbind(pbmc_iDAcellembedding, pbmc[["RNA"]]@counts["MS4A1",], pbmc_iDA_ident))

#markers GNLY/NKG7: NK cells
GNLYiDA = as.data.frame(cbind(pbmc_iDAcellembedding, pbmc[["RNA"]]@counts["GNLY",], pbmc_iDA_ident))
NKG7iDA = as.data.frame(cbind(pbmc_iDAcellembedding, pbmc[["RNA"]]@counts["NKG7",], pbmc_iDA_ident))

#marker PPBP: Dendritic Cells
PPBPiDA = as.data.frame(cbind(pbmc_iDAcellembedding, pbmc[["RNA"]]@counts["PPBP",], pbmc_iDA_ident))

```


## plot markers versus LD weight
```{r, include = FALSE}
#monocytes

  ggplot(data = CD14iDA, aes(x = pbmc_iDAcellembedding[,6], y = pbmc[["RNA"]]@data["CD14",])) + 
   geom_point(aes(color = as.factor(pbmc_iDA_ident))) + 
   theme_classic() +
   xlab("LD6 Weight") + 
   ylab("CD14 Expression") + 
   labs(color = "iDA Cluster Identity")
 
 ggplot(data = LYZiDA, aes(x = pbmc_iDAcellembedding[,6], y = pbmc[["RNA"]]@data["LYZ",])) + 
   geom_point(aes(color = as.factor(pbmc_iDA_ident))) + 
   theme_classic() + 
   xlab("LD6 Weight") + 
   ylab("LYZ Expression") + 
   labs(color = "iDA Cluster Identity")
 
#NK cells
 ggplot(data = GNLYiDA, aes(x = pbmc_iDAcellembedding[,3], y = pbmc[["RNA"]]@data["GNLY",])) + 
   geom_point(aes(color = as.factor(pbmc_iDA_ident))) + 
   xlab("LD3 Weight") + 
   ylab("GNLY Expression")  + 
   labs(color = "iDA Cluster Identity") + 
   theme_classic() 

 
 ggplot(data = NKG7iDA, aes(x = pbmc_iDAcellembedding[,3], y = pbmc[["RNA"]]@data["NKG7",])) + 
   geom_point(aes(color = as.factor(pbmc_iDA_ident))) + 
   xlab("LD3 Weight") + 
   ylab("NKG7 Expression") + 
   labs(color = "iDA Cluster Identity") + 
   theme_classic()
 
# #B-cells
 ggplot(data = MS4A1iDA, aes(x = pbmc_iDAcellembedding[,5], y = pbmc[["RNA"]]@data["MS4A1",])) + 
   geom_point(aes(color = as.factor(pbmc_iDA_ident))) +  
   theme_classic() +
   xlab("LD5 Weight") + 
   ylab("MS4A1 Expression") + 
   labs(color = "iDA Cluster Identity") 
# 
# #Platelets
 ggplot(data = PPBPiDA, aes(x = pbmc_iDAcellembedding[,1], y = pbmc[["RNA"]]@data["PPBP",])) + 
   geom_point(aes(color = as.factor(pbmc_iDA_ident))) + 
   theme_classic() + 
   labs(color = "iDA Cluster Identity") + 
   xlab("LD1 Weight") + 
   ylab("PPBP Expression")
 
```


## difference in scaled expression for cell types for each marker
```{r}
#cluster 1 = monocytes
pbmc_CD14 = pbmc[["RNA"]]@scale.data["CD14",]
  mean(pbmc_CD14[pbmc_iDA_ident == 1])
  mean(pbmc_CD14[pbmc_iDA_ident != 1])

pbmc_LYZ = pbmc[["RNA"]]@scale.data["LYZ",]
  mean(pbmc_LYZ[pbmc_iDA_ident == 1])
  mean(pbmc_LYZ[pbmc_iDA_ident != 1])

  
#cluster 8 = NK cells
pbmc_GNLY = pbmc[["RNA"]]@scale.data["GNLY",]
  mean(pbmc_GNLY[pbmc_iDA_ident == 6])
  mean(pbmc_GNLY[pbmc_iDA_ident != 6])

pbmc_NKG7 = pbmc[["RNA"]]@scale.data["NKG7",]
  mean(pbmc_NKG7[pbmc_iDA_ident == 6])
  mean(pbmc_NKG7[pbmc_iDA_ident != 6])


#cluster 5 = bcells
pbmc_MS4A1 = pbmc[["RNA"]]@scale.data["MS4A1",]
  mean(pbmc_MS4A1[pbmc_iDA_ident == 5])
  mean(pbmc_MS4A1[pbmc_iDA_ident != 5])
  

#cluster 10 = dendritic cells
pbmc_PPBP = pbmc[["RNA"]]@scale.data["PPBP",]
  mean(pbmc_PPBP[pbmc_iDA_ident == 9])
  mean(pbmc_PPBP[pbmc_iDA_ident != 9])
```

# Dunn index
```{r}
cellbygenePBMC = t(pbmc[["RNA"]]@scale.data)

#distance matrix (takes awhile to compute)
Dist <- dist(cellbygenePBMC,method="euclidean")

#Seurat clusters
pbmc <- FindNeighbors(pbmc, dims = 1:10)
pbmc <- FindClusters(pbmc)
seurat_clusters <- Idents(pbmc)

#iDA clusters
iDAPBMC_clusters <- pbmc@meta.data$iDA_clust
names(iDAPBMC_clusters) <- colnames(pbmc[["RNA"]]@counts)

seuratclusts_Dunn = dunn(Dist, as.numeric(seurat_clusters))
iDAclustsPBMC_Dunn = dunn(Dist, as.numeric(iDAPBMC_clusters))

confusionMatrix(as.factor(iDAPBMC_clusters), reference = as.factor(as.numeric(seurat_clusters)))

DIs = as.data.frame(c(seuratclusts_Dunn, iDAclustsPBMC_Dunn))
rownames(DIs) <- c("Seurat Clusters", "iDA Clusters")
colnames(DIs) <- "Dunn Index"

DIs
```



# tSNE of iDA versus PCA embeddings
```{r}
#tSNE for PCA and iDA
#run tSNE for each reduction
  set.seed(10)
  tsne_pca <- Rtsne(pbmc[["pca"]]@cell.embeddings[,1:10])
  tsne_iDA <- Rtsne(pbmc[["iDA"]]@cell.embeddings[,1:8])

  #tSNE plots for each reduction
    ggplot(as.data.frame(tsne_pca$Y), aes(x = tsne_pca$Y[,1], y = tsne_pca$Y[,2], color = as.factor(Idents(pbmc)))) +
      geom_point() + labs(color = "Seurat Clusters") + xlab("tSNE 1") + ylab("tSNE 2") + ggtitle("tSNE of PCA cell embeddings") +
      theme_classic()
    ggplot(as.data.frame(tsne_iDA$Y), aes(x = tsne_iDA$Y[,1], y = tsne_iDA$Y[,2], color = as.factor(pbmc_iDA_ident))) + 
      geom_point() + labs(color = "iDA Clusters") + xlab("tSNE 1") + ylab("tSNE 2") + ggtitle("tSNE of iDA cell embeddings") +
      theme_classic()
```

# UMAP of iDA versus PCA embeddings
```{r}
#UMAP comparison of PCA versus iDA embedding
  #run UMAP for each reduction
    set.seed(10)
    UMAPpcapbmc <- umap(pbmc[["pca"]]@cell.embeddings[,1:10])
    
    set.seed(10)
    UMAPiDApbmc <- umap(pbmc[["iDA"]]@cell.embeddings[,1:8])
   
  #UMAP plots for each reduction
    ggplot(as.data.frame(UMAPpcapbmc$layout), aes(x =UMAPpcapbmc$layout[,1], y = UMAPpcapbmc$layout[,2], color = as.factor(Idents(pbmc)))) +
      geom_point() + labs(color = "Seurat Clusters") + xlab("UMAP 1") + ylab("UMAP 2") + ggtitle("UMAP of PCA cell embeddings") +
      theme_classic()
    
      ggplot(as.data.frame(UMAPiDApbmc$layout), aes(x =UMAPiDApbmc$layout[,1], y = UMAPiDApbmc$layout[,2], color = as.factor(Idents(pbmc)))) +
      geom_point() + labs(color = "iDA Clusters") + xlab("UMAP 1") + ylab("UMAP 2") + ggtitle("UMAP of iDA cell embeddings") +
      theme_classic()
```

