---
title: "sce H1975 heterogenerity: 10x 5 Clusters"
author: "Theresa Alexander"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: yes
  html_document:
    number_sections: yes
    toc: yes
---

# Load Data
The data can be downloaded from the sc_mixology repo here:

https://github.com/LuyiTian/sc_mixology/blob/master/data/sincell_with_class.RData
https://github.com/LuyiTian/sc_mixology/blob/master/data/sincell_with_class_5cl.RData

```{r, message = FALSE}
load("~/Desktop/iDA_bioc/data/sincell_with_class.RData")
load("~/Desktop/iDA_bioc/data/sincell_with_class_5cl.RData")
```

# Packages
```{r}
library("scran")
library("scater")
library("igraph")
library("Rtsne")
library("stats")
library("mclust")
library("ggplot2")
library("gridExtra")
library("irlba")
library("MASS")
library("klaR")
library("dplyr")
```

# Helper Functions
```{r, include = FALSE}
getSNN <- function(data.use,
                   k.param = 10,
                   prune.SNN = 1/15,
                   set.seed = FALSE) {
  data.use <- as.matrix(data.use)
  n.obs <- nrow(x = data.use)
  
  if (n.obs < k.param) {
    warning(
      "k.param set larger than number of cells. Setting k.param to number of cells - 1.",
      call. = FALSE
    )
    k.param <- n.obs - 1
  }
  
  ## TODO: refactor this to avoid code duplication
  if (!is.numeric(set.seed)){
    
    SNN_igraph <- scran::buildKNNGraph(
      data.use,
      k = k.param,
      transposed = TRUE)
    snn.matrix <- similarity(
      SNN_igraph,
      method = "jaccard")
    
    snn.matrix[snn.matrix < 1/15] <- 0
    rownames(x = snn.matrix) <- rownames(x = data.use)
    colnames(x = snn.matrix) <- rownames(x = data.use)
    
    snn.graph <- graph_from_adjacency_matrix(snn.matrix, weighted = TRUE, mode = "undirected")
    return(snn.graph)
    
  } else if (is.numeric(set.seed)){
    set.seed(set.seed)
    
    SNN_igraph <- scran::buildKNNGraph(
      data.use,
      k = k.param,
      transposed = TRUE)
    snn.matrix <- similarity(
      SNN_igraph,
      method = "jaccard")
    
    snn.matrix[snn.matrix < 1/15] <- 0
    rownames(x = snn.matrix) <- rownames(x = data.use)
    colnames(x = snn.matrix) <- rownames(x = data.use)
    
    snn.graph <- graph_from_adjacency_matrix(snn.matrix, weighted = TRUE, mode = "undirected")
    return(snn.graph)
  }
}

get_walktrapClusters <- function(embedding, c.param = NA){  
  snn <- getSNN(data.use = embedding, set.seed = 11)
  set.seed(11)
  if (!is.na(c.param)){
  walktrapClusters <- suppressWarnings(igraph::cluster_walktrap(snn))
  maxmodclust <- igraph::cut_at(walktrapClusters, n = c.param)
  } else {
      modularity <- c(0)
      for (i in 2:15){
        walktrapClusters <- suppressWarnings(igraph::cluster_walktrap(snn))
        modularity <- c(modularity,  modularity(snn, suppressWarnings(igraph::cut_at(walktrapClusters, n = i))))
      }
      maxmodclust <- igraph::cut_at(walktrapClusters, n = which.max(modularity)+1)
  }
  return(maxmodclust)
}


CumulativeARI <- function(embedding, knownlabels, c.param = NA){
  ARIs <- c()
   for (i in 1:(dim(embedding)[2])) {
     walktrapClusters <- get_walktrapClusters(embedding[,1:i], c.param)
     ARI <- adjustedRandIndex(walktrapClusters, knownlabels)
    ARIs <- c(ARIs, ARI) 
       #pick highest modularity 
    }
  return(ARIs)
}




CumulativeFstatistic <- function(embedding, labels) {
  RSS <- c()
  i = 1
    for (i in 1:(dim(embedding)[2])) {
      RSS[i] = anova(lm(as.numeric(embedding[,i]) ~
                             as.factor(labels)))[2,2]
    }
  
  TSS = c()
  for (i in 1:(dim(embedding)[2])) {
    TSS[i] = sum(anova(lm(as.numeric(embedding[,i]) ~
                               as.factor(labels)))[,2])
  }
  
  Fstat <- c()
  i = 1
  for (i in 1:length(TSS)){
    Fstat <- c(Fstat, (sum(TSS[1:i]) - sum(RSS[1:i]))/sum(RSS[1:i]))
  }
  return(Fstat)
}


```


# 3 Cell Types (10x_3cl)
```{r}
#scran normalize
sce_sc_10x_qc <- computeSumFactors(sce_sc_10x_qc)
sce_sc_10x_qc <- logNormCounts(sce_sc_10x_qc)

#get variable genes
stats <- scran::modelGeneVar(logcounts(sce_sc_10x_qc))
var.features <- scran::getTopHVGs(stats, n = 3000)

#PCA reduction
sce_sc_10x_qc <- runPCA(sce_sc_10x_qc, subset_row = var.features)


#tsne
set.seed(1234)
sce10x_tsne <- Rtsne(reducedDim(sce_sc_10x_qc,"PCA")[,1:10])

Cell_line <- sce_sc_10x_qc$cell_line_demuxlet

#plot tsne dims
ggplot(as.data.frame(sce10x_tsne$Y), aes(x = sce10x_tsne$Y[,1], y = sce10x_tsne$Y[,2], col=Cell_line)) + 
  geom_point() +
  theme_classic() +
  ggtitle("sc_10x 3 Cell Lines") +
  xlab("TSNE 1") +
  ylab("TSNE 2") 
```




## Cluster PCA
```{r}
PCAclusters <- get_walktrapClusters(reducedDim(sce_sc_10x_qc,"PCA")[,1:10])
names(PCAclusters) <- colnames(sce_sc_10x_qc)

PCAclusters_c3 <- get_walktrapClusters(reducedDim(sce_sc_10x_qc,"PCA")[,1:10], c.param = 3)
names(PCAclusters_c3) <- colnames(sce_sc_10x_qc)
```


## iDA on 10X 3cl
### unsupervised # of clusters
```{r}
TENx_iDA <- iDA_core(data.use = logcounts(sce_sc_10x_qc), NormCounts = logcounts(sce_sc_10x_qc), set.seed = 11, diag=TRUE)

TENX_3cl_iDAclusters <- as.factor(TENx_iDA[[1]])
```

### supervised # of clusters
```{r}
TENx_iDA_3cl <- iDA_core(logcounts(sce_sc_10x_qc), NormCounts = logcounts(sce_sc_10x_qc), set.seed = 11, diag=TRUE, c.param = 3)

TENX_3cl_iDAclusters_3cl <- as.factor(TENx_iDA_3cl[[1]])
```


## Plot the 10X 3cl TSNE with iDA cluster colors

```{r}
`3 Cell Lines` <- sce_sc_10x_qc$cell_line
g1 <- ggplot(as.data.frame(sce10x_tsne$Y), aes(x = sce10x_tsne$Y[,1], y = sce10x_tsne$Y[,2], col=`3 Cell Lines`)) + geom_point() +
  theme_classic() +
  ggtitle("3 Cell Line Mix") +
  xlab("tSNE 1") +
  ylab("tSNE 2") +  
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10)) 

g2 <- ggplot(as.data.frame(sce10x_tsne$Y), aes(x = sce10x_tsne$Y[,1], y = sce10x_tsne$Y[,2], col=TENX_3cl_iDAclusters_3cl)) + 
  geom_point() +
  theme_classic() +
  ggtitle("tSNE of PCA reduced sc_10x 3 clusters, colors: iDA, c.param = 3") +
  xlab("TSNE 1") +
  ylab("TSNE 2") 

g3 <- ggplot(as.data.frame(sce10x_tsne$Y), aes(x = sce10x_tsne$Y[,1], y = sce10x_tsne$Y[,2], col=Cell_line)) + 
  geom_point() +
  theme_classic() +
  ggtitle("tSNE of PCA reducted sc_10x 3 clusters, colors: Known Cell Line") +
  xlab("TSNE 1") +
  ylab("TSNE 2") 


 PCAclusters_supervised <- as.factor(PCAclusters_c3)
g4 <- ggplot(as.data.frame(sce10x_tsne$Y), aes(x = sce10x_tsne$Y[,1], y = sce10x_tsne$Y[,2], col=PCAclusters_supervised)) + 
  geom_point() +
  theme_classic() +
  ggtitle("tSNE of PCA reducted sc_10x 3 clusters, colors: PCA + walktrap clusters, c.param = 3") +
  xlab("TSNE 1") +
  ylab("TSNE 2") 


PCAclusters_unsup <- as.factor(PCAclusters)
g5 <- ggplot(as.data.frame(sce10x_tsne$Y), aes(x = sce10x_tsne$Y[,1], y = sce10x_tsne$Y[,2], col=PCAclusters_unsup)) + 
  geom_point() +
  theme_classic() +
  ggtitle("tSNE of PCA reduced sc_10x 3 clusters, colors: PCA + walktrap clsuters, unsupervised") +
  xlab("TSNE 1") +
  ylab("TSNE 2") 


g1
```


## Cumulative Fstat
```{r}
#unsupervised iDA
TENx_iDA_CumulativeFstat <- CumulativeFstatistic(embedding = TENx_iDA[[2]], labels = colData(sce_sc_10x_qc)$cell_line)

#supervised iDA
TENx_iDA_3cl_CumulativeFstat <- CumulativeFstatistic(embedding = TENx_iDA_3cl[[2]], labels = colData(sce_sc_10x_qc)$cell_line)


#PCA
TENx_PCA_3cl_CumulativeFstat <- CumulativeFstatistic(embedding = reducedDim(sce_sc_10x_qc,"PCA")[,1:10], labels = colData(sce_sc_10x_qc)$cell_line)

CumulativeFstat_10x3cl <- as.data.frame(cbind("iDAunsup" = as.numeric(c(TENx_iDA_CumulativeFstat, NA)), "iDAsup" = c(TENx_iDA_3cl_CumulativeFstat, NA, NA, NA, NA, NA, NA, NA, NA), "PCA" = TENx_PCA_3cl_CumulativeFstat))

cl3Fstat <- ggplot(CumulativeFstat_10x3cl, aes(x = c(1:10), y = iDAunsup)) + 
  geom_point(aes(x = c(1:10), y = iDAunsup, color = "iDA")) +
  geom_path(aes(x = c(1:10), y = iDAunsup, color = "iDA")) +
  geom_point(aes(x = c(1:10), y = iDAsup, color = "iDA Supervised")) +
  geom_path(aes(x = c(1:10), y = iDAsup, color = "iDA Supervised")) +
  geom_point(aes(x = c(1:10), y = PCA, color = "PCA")) +
  geom_path(aes(x = c(1:10), y = PCA, color = "PCA")) +
  xlab("Dimension") + 
  ylab("Cumulative F-statistic") +
  ggtitle("Cumulative F-statistic", subtitle = "3 Cell Lines") +
  theme_classic() +  
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10)) 

cl3Fstat
```














# 5 Cell Types (10x_5cl)
#scran normalization and PCA
```{r}
#scran normalize
sce_sc_10x_5cl_qc <- computeSumFactors(sce_sc_10x_5cl_qc)
sce_sc_10x_5cl_qc <- logNormCounts(sce_sc_10x_5cl_qc)

#get variable genes
stats <- scran::modelGeneVar(logcounts(sce_sc_10x_5cl_qc))
var.features <- scran::getTopHVGs(stats, n = 3000)

#PCA reduction
sce_sc_10x_5cl_qc <- runPCA(sce_sc_10x_5cl_qc)


#tsne
set.seed(123)
sce10x_5cl_tsne <- Rtsne(reducedDim(sce_sc_10x_5cl_qc,"PCA")[,1:10])

`5 Cell Lines` <- sce_sc_10x_5cl_qc$cell_line_demuxlet

#plot tsne dims
Fivecellmix <- ggplot(as.data.frame(sce10x_5cl_tsne$Y), aes(x = sce10x_5cl_tsne$Y[,1], y = sce10x_5cl_tsne$Y[,2], col=`5 Cell Lines`)) + 
  geom_point() +
  theme_classic() +
  ggtitle("5 Cell Line Mix") +
  xlab("tSNE 1") +
  ylab("tSNE 2") +  
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10)) 

Fivecellmix
```



## Cluster PCA
```{r}
PCAclusters_5cl <- get_walktrapClusters(reducedDim(sce_sc_10x_5cl_qc,"PCA")[,1:10])
names(PCAclusters_5cl) <- colnames(sce_sc_10x_5cl_qc)

PCAclusters_c5 <- get_walktrapClusters(reducedDim(sce_sc_10x_5cl_qc,"PCA")[,1:10], c.param = 5)
names(PCAclusters_c5) <- colnames(sce_sc_10x_5cl_qc)
```



# iDA on 10X 5cl
## unsupervised # of clusters
```{r}
TENx_iDA_5cl <- iDA(logcounts(sce_sc_10x_5cl_qc), NormCounts = logcounts(sce_sc_10x_5cl_qc), set.seed = 11)
TENX_5cl_iDAclusters <- as.factor(TENx_iDA_5cl[[1]])
```

## supervised # of clusters
```{r}
TENx_iDA_5cl_sup <- iDA(logcounts(sce_sc_10x_5cl_qc), NormCounts = logcounts(sce_sc_10x_5cl_qc), set.seed = 11, c.param = 5)

TENX_3cl_iDAclusters_5cl <- as.factor(TENx_iDA_5cl_sup[[1]])
```

## Cumulative Fstat
```{r}
#unsupervised iDA
TENx_iDA5cl_CumulativeFstat_unsup <- CumulativeFstatistic(embedding = TENx_iDA_5cl[[2]], labels = colData(sce_sc_10x_5cl_qc)$cell_line_demuxlet)

#supervised iDA
TENx_iDA_5cl_CumulativeFstat_sup <- CumulativeFstatistic(embedding = TENx_iDA_5cl_sup[[2]], labels = colData(sce_sc_10x_5cl_qc)$cell_line_demuxlet)

#PCA
TENx_PCA_5cl_CumulativeFstat <- CumulativeFstatistic(embedding = reducedDim(sce_sc_10x_5cl_qc,"PCA")[,1:13], labels = colData(sce_sc_10x_5cl_qc)$cell_line_demuxlet)

CumulativeFstat_10x5cl <- as.data.frame(cbind("iDAunsup" = as.numeric(c(TENx_iDA5cl_CumulativeFstat_unsup)), "iDAsup" = c(TENx_iDA_5cl_CumulativeFstat_sup, NA, NA, NA, NA, NA, NA, NA, NA, NA), "PCA" = TENx_PCA_5cl_CumulativeFstat))

Fstat5CL <- ggplot(CumulativeFstat_10x5cl, aes(x = c(1:13), y = iDAunsup)) + 
  geom_point(aes(x = c(1:13), y = iDAunsup, color = "iDA")) +
  geom_path(aes(x = c(1:13), y = iDAunsup, color = "iDA")) +
  geom_point(aes(x = c(1:13), y = iDAsup, color = "iDA Supervised")) +
  geom_path(aes(x = c(1:13), y = iDAsup, color = "iDA Supervised")) +
  geom_point(aes(x = c(1:13), y = PCA, color = "PCA")) +
  geom_path(aes(x = c(1:13), y = PCA, color = "PCA")) +
  xlab("Dimension") + 
  ylab("Cumulative F-statistic") +
  ggtitle("Cumulative F-statistic", subtitle = "5 Cell Lines") +
  theme_classic() +  
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10), legend.position = "none") 

```
